---
title: "Bellabeat Case Study"
author: "Ahmad Abdelbaset"
date: "2024-07-06"
output: html_document
---
## Bellabeat Project (Google Data Analytics Capstone)

We start by installing the required packaes


``` {r }
install.packages("tidyverse")
install.packages("here")
install.packages("skimr")
install.packages("janitor")
install.packages("lubridate")
install.packages("readr")

```

Then, we will bring the libraries:
```{r}
library(tidyverse)
library(here)
library(skimr)
library(janitor)
library(lubridate)
library(readr)
```

and then we will bring the files of dataset:
```{r}
dailyActivity <- read.csv("dailyActivity_merged.csv", header = TRUE)
heartrate <- read.csv("heartrate_seconds_merged.csv", header = TRUE)
sleepDay <- read.csv("sleepDay_merged.csv", header = TRUE)
hourlyCalories <- read.csv("hourlyCalories_merged.csv", header = TRUE)
minuteSleep <- read.csv("minuteSleep_merged.csv", header = TRUE)
sleepDay <- read.csv("sleepDay_merged.csv", header = TRUE)
weight <- read.csv("weightLogInfo_merged.csv", header = TRUE)
```

Lets see the data

```{r}
head(dailyActivity)
head(heartrate)
head(sleepDay)
head(hourlyCalories)
head(minuteSleep)
head(weight)
```

Before starting analyzing, we need to take a look on the structure of data:

```{r}
str(dailyActivity)
str(heartrate)
str(sleepDay)
str(hourlyCalories)
str(minuteSleep)
str(weight)
```


We need to ensure about the number of participant in each file:

```{r}
n_unique(dailyActivity$Id)
n_unique(heartrate$Id)
n_unique(sleepDay$Id)
n_unique(hourlyCalories$Id)
n_unique(minuteSleep$Id)
n_unique(weight$Id)
```

All data sets have 33 participants each except the heartrate, and weight data set which has 7, and 8 participants respectively.
I will drop the heartrate and weight data sets because 7 and 8 participants is too small a sample to draw conclusions and make recommendations.

Lets check for Duplicates:

```{r}
sum(duplicated(dailyActivity))
sum(duplicated(sleepDay))
sum(duplicated(hourlyCalories))
sum(duplicated(minuteSleep))
sum(duplicated(sleepDay))

```
Then we have duplicates, we have to remove it:

```{r}
dailyActivity <- dailyActivity %>%
  distinct() %>%
  drop_na()
sleepDay <- sleepDay %>%
  distinct() %>%
  drop_na()
hourlyCalories <- hourlyCalories %>%
  distinct() %>%
  drop_na()
minuteSleep <- minuteSleep %>%
  distinct() %>%
  drop_na()

sleepDay <- sleepDay %>%
  distinct() %>%
  drop_na()

```

Make date and time columns consistent
```{r}
dailyActivity <- dailyActivity %>%
  rename(date = ActivityDate) %>%
  mutate(date = as_date(date, format = "%m/%d/%Y"))

sleepDay <- sleepDay %>%
  rename(date = SleepDay) %>%
  mutate(date = as_date(date, format ="%m/%d/%Y %I:%M:%S %p", tz = Sys.timezone()))

hourlyCalories <- hourlyCalories %>% 
  rename(date_time = ActivityHour) %>% 
  mutate(date_time = as.POSIXct(date_time, format ="%m/%d/%Y %I:%M:%S %p" , tz=Sys.timezone()))

minuteSleep <- minuteSleep %>% 
  rename(date_time = date) %>% 
  mutate(date_time = as.POSIXct(date, format ="%m/%d/%Y %I:%M:%S %p" , tz=Sys.timezone()))

sleepDay <- sleepDay %>% 
  rename(date = date) %>%
  mutate(date = as_date(date, format = "%m/%d/%Y"))

```

Now, I will merge dailyActivity and sleepDay:
```{r}
daily_activity_sleep <- merge(dailyActivity, sleepDay, by= c("Id", "date"))
```

I will subtract the TotalMinutesASleep from TotalTimeInBed to know the pure time of seelping and to see how long people are stay in the bed without sleeping i will call for this "InBedWithoutSleeping"

```{r}
daily_activity_sleep$InBedWithoutSleeping <- daily_activity_sleep$TotalTimeInBed - daily_activity_sleep$TotalMinutesAsleep

```

```{r}
daily_average <- daily_activity_sleep %>%
  group_by(Id) %>%
  summarise (mean_daily_steps = mean(TotalSteps),   mean_daily_calories = mean(Calories), mean_daily_sleep = mean(TotalMinutesAsleep), mean_time_inBed_without_sleep=mean(InBedWithoutSleeping))
```

Now I want to calssiy the type of users into four groups according to their average total steps:
Sedentary: fewer than 5,000 steps per day
Lightly_Active: between 5,000 and 7,499 steps per day
Fairly_Active: between 7,500 and 9,999 steps per day
Highly_Active: 10,000 steps per day or more

```{r}
classify_activity <- function(steps) {
  if (steps < 5000) {
    return("Sedentary")
  } else if (steps >= 5000 & steps < 7500) {
    return("Lightly_Active")
  } else if (steps >= 7500 & steps < 10000) {
    return("Fairly_Active")
  } else {
    return("Highly_Active")
  }
}

daily_average$user_type <- daily_average %>%
  mutate(user_type = sapply(mean_daily_steps, classify_activity))
```

Lets see the distribution of user_types:
```{r}
freq <- table(daily_average$user_type$user_type)
percentages <- prop.table(freq) * 100

# Create the pie chart with percentages
pie(freq, labels = paste(names(freq), ": ", freq, " (", round(percentages, 1), "%)", sep = ""), main = "Distribution of User Types")


```


Now I will create a data frame that shows the average calories per hour during 24 hours, the file will be used is (hourlyCalories):

```{r}
AverageCaloriesPerHour <- hourlyCalories %>%
    mutate(hour = format(date_time, "%H")) %>%
    group_by(hour) %>%
    summarise(avg_calories = mean(Calories, na.rm = TRUE))

```

Lets visualize it and see what in which hours people loss more calories:

```{r}
ggplot(AverageCaloriesPerHour, aes(x = hour, y = avg_calories)) + geom_bar(stat = "identity", fill = "steelblue") + labs(title = "Average Calories", x = "Hour", y = "Calories") +
  theme_minimal()

```

Lets see now the how the average sleep minutes and average steps in each weekday:

```{r}

weekday_steps_sleep <- daily_activity_sleep %>%
  mutate(weekday = weekdays(date))

weekday_steps_sleep$weekday <-ordered(weekday_steps_sleep$weekday, levels=c("Monday", "Tuesday", "Wednesday", "Thursday",
"Friday", "Saturday", "Sunday"))

 weekday_steps_sleep <-weekday_steps_sleep%>%
  group_by(weekday) %>%
  summarize (mean_steps=mean(TotalSteps), mean_sleep = mean(TotalMinutesAsleep))
```

Lets plot it and see how they sleep and walk in weekdays:
```{R}
install.packages("ggpubr")

library(ggpubr)
```

```{R}
ggarrange(
    ggplot(weekday_steps_sleep) +
      geom_col(aes(weekday, mean_steps), fill = "#006699") +
      geom_hline(yintercept = 7500) +
      labs(title = "Daily steps per weekday", x= "", y = "") +
      theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 1)), ggplot(weekday_steps_sleep, aes(weekday, mean_sleep)) +
      geom_col(fill = "#85e0e0") +
      geom_hline(yintercept = 480) +
      labs(title = "Minutes asleep per weekday", x= "", y = "") +
      theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 1))
  )
```


Now, lets see how the users are using their smart device:

```{r}

daily_use <- daily_activity_sleep %>%
  group_by(Id) %>%
  summarize(days_used=sum(n())) %>%
  mutate(usage = case_when(
    days_used >= 1 & days_used <= 10 ~ "low use",
    days_used >= 11 & days_used <= 20 ~ "moderate use", 
    days_used >= 21 & days_used <= 31 ~ "high use", 
  ))
  
head(daily_use)
```

We will see the percentage and then visualize it:

```{r}

daily_use_percent <- daily_use %>%
  group_by(usage) %>%
  summarise(total = n()) %>%
  mutate(totals = sum(total)) %>%
  group_by(usage) %>%
  summarise(total_percent = total / totals) %>%
  mutate(labels = scales::percent(total_percent))

daily_use_percent$usage <- factor(daily_use_percent$usage, levels = c("high use", "moderate use", "low use"))

head(daily_use_percent)
```


We classified the users according to their usage time, and now we have to visualize the result:

```{r}
daily_use_percent %>%
  ggplot(aes(x="",y=total_percent, fill=usage)) +
  geom_bar(stat = "identity", width = 1)+
  coord_polar("y", start=0)+
  theme_minimal()+
  theme(axis.title.x= element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(), 
        panel.grid = element_blank(), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size=14, face = "bold")) +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5))+
  scale_fill_manual(values = c("#006633","#00e673","#80ffbf"),
                    labels = c("High use - 21 to 31 days",
                                 "Moderate use - 11 to 20 days",
                                 "Low use - 1 to 10 days"))+
  labs(title="Daily use of smart device")

```
Now, Lets see the how the correlation looks like between (daily steps vs calories) and (daily steps vs sleep)

```{r}
ggplot(data= subset(daily_activity_sleep,!is.na(TotalMinutesAsleep)),aes(TotalSteps,TotalMinutesAsleep))+
    geom_rug(position= "jitter", size=.08)+
    geom_jitter(alpha= 0.5)+
    geom_smooth(color= "blue", linewidth=.6)+
    stat_cor(method = "pearson", label.x = 15000, label.y = 650)+
    labs(title= "Daily steps vs. sleep", x= "Daily Steps", y= "Minutes Asleep")+
    theme_minimal()

```
```{r}
ggplot(daily_activity_sleep,aes(TotalSteps,Calories))+geom_jitter(alpha=.5)+
    geom_rug(position="jitter", linewidth=.08)+
    geom_smooth(linewidth =.6)+
    stat_cor(method = "pearson", label.x = 20000, label.y = 2300)+
    labs(title= "Daily Steps vs. Calories", x= "Daily steps", y="Calories")+
    theme_minimal()

```

